---
title: "Running Back Straight-Line Speed"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
library(rdtools)
library(tidyverse)
library(ggthemes)
library(nflfastR)
library(mltools)
library(data.table)
library(xgboost)
library(caret)
library(vip)
library(SHAPforxgboost)
library(ggimage)

theme_reach <- function() {
  theme_fivethirtyeight() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 18, hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(size = 14, hjust = 0.5),
      axis.title.x = element_text(size=14),
      axis.title.y = element_text(size=14),
      axis.text = element_text(size = 12),
      strip.text = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12)
    )
}
```

## Loading in the data

Using tracking data, we were able to get a running back's speed from the time of handoff to the time of first contact.
Doing this allowed us to measure their straight-line speed to see their explsoivness. We then put the data into a play sheet and are able to load that in. 

```{r loading in data, echo = TRUE}
rushing_data <<- pull_s3(paste0("analytics/projections/by_facet/", 'nfl', "/%i/rushing.csv.gz"), season_start = 2017, season_end = 2020)

play_speed_all_18 <- read.csv("~/tracking-pff/play-speed-year/play_speed_all_18.csv")
play_speed_all_19 <- read.csv("~/tracking-pff/play-speed-year/play_speed_all_19.csv")
play_speed_all_20 <- read.csv("~/tracking-pff/play-speed-year/play_speed_all_20.csv")

play_speed_all <- rbind(play_speed_all_18, play_speed_all_19, play_speed_all_20)

play_speed_all <- play_speed_all %>%
  dplyr::select(gameid, playid, player, avg_speed, seconds_before_contact, ybc)
```

```{r chaning rushing data, echo = FALSE}
rushing_data <- rushing_data %>%
  mutate(offense = case_when(
    offense == "SD" ~ "LAC",
    offense == "BLT" ~ "BAL",
    offense == "OAK" ~ "LV",
    offense == "HST" ~ "HOU",
    offense == "SL" ~ "LA",
    offense == "CLV" ~ "CLE", 
    offense == "ARZ" ~ "ARI",
    TRUE ~ offense
  )) %>%
  ungroup()

rushing_data <- rushing_data %>%
  mutate(defense = case_when(
    defense == "SD" ~ "LAC",
    defense == "BLT" ~ "BAL",
    defense == "OAK" ~ "LV",
    defense == "HST" ~ "HOU",
    defense == "SL" ~ "LA",
    defense == "CLV" ~ "CLE", 
    defense == "ARZ" ~ "ARI",
    TRUE ~ defense
  )) %>%
  ungroup()

rushing_data <- rushing_data %>%
  mutate(yards_before_contact = yards - yards_after_contact)

rushers <- rushing_data %>%
  dplyr::select(game_id, play_id, player, yards_before_contact)

rb_colors <- rushing_data %>%
  group_by(player, offense) %>%
  summarize(plays = n()) %>%
  arrange(-plays) %>%
  group_by(player) %>%
  top_n(n = 1) %>%
  left_join(teams_colors_logos, by = c("offense" = "team_abbr"))
```

## Exploratory data analysis

```{r linear regression}
round(summary(lm(ybc ~ avg_speed, data = play_speed_all))$r.squared, 2)
round(summary(lm(ybc ~ seconds_before_contact, data = play_speed_all))$r.squared, 2)
summary(lm(ybc ~ (avg_speed + seconds_before_contact)^2, data = play_speed_all))
```

We can look at the plots of these:

```{r plot-1, echo=FALSE, fig.align = "center", out.width= "100%"}
play_speed_all %>%
  ggplot(aes(x = avg_speed, y = ybc)) +
  geom_jitter(alpha = 0.5, size = 1.5) +
  geom_smooth(size = 2, color = "darkorange") +
  theme_reach() +
  labs(x = "Average Speed",
       y = "Yards Before Contact",
       title = "Higher Speed Before Contact Leads \n to Higher Yards Before Contact On a Play",
       subtitle = "2017-2020, Weeks 1-17") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  annotate("text", x = 3, y = 50, label = "R^2 = 0.15", size = 5)
```

The greater effect is the amount of seconds before contact but both are meaningful for determining yards before contact.

```{r plot-2, echo=FALSE, fig.align = "center", out.width= "100%"}
play_speed_all %>%
  filter(seconds_before_contact <= 8) %>%
  ggplot(aes(x = seconds_before_contact, y = ybc)) +
  geom_jitter(alpha = 0.5, size = 2) +
  geom_smooth(size = 2, color = "darkgreen") +
  theme_reach() +
  labs(x = "Seconds Before Contact",
       y = "Yards Before Contact",
       title = "Higher Seconds Before Contact Leads to \n Higher Yards Before Contact On a Play",
       subtitle = "2017-2020, Weeks 1-17") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  annotate("text", x = 3, y = 50, label = "R^2 = 0.31", size = 5)
```

## What Influences Speed

Not all speeds are built the same. Based on down, distance, box defenders, yardline and run scheme speed can change for the average running back. 

First we look at down and distance. 

```{r plot 3, echo = FALSE, fig.align = "center", out.width= "100%"}
rushing_data_speed <- rushing_data %>%
  dplyr::select(game_id, play_id, offense, defense, season, week, down, distance, player_id,
                position, quarter, seconds_left_in_quarter, concept_1, concept_2, run_position,
                intended_run_position, run_direction, box_players, rpo, yards,
                yards_after_contact, avoided_tackles, yards_to_go) %>%
  left_join(play_speed_all, by = c("game_id" = "gameid", "play_id" = "playid")) %>%
  filter(!is.na(avg_speed)) %>%
  dplyr::filter(grepl("HB",position))

rushing_data_speed$avoided_tackles[is.na(rushing_data_speed$avoided_tackles)] <- 0

rushing_data_speed %>%
  filter(distance < 15) %>%
  filter(down %in% c(1, 2, 3)) %>%
  group_by(down, distance) %>%
  summarize(rushes = n(),
            avg_speed = mean(avg_speed)) %>%
  mutate(Down = as.factor(down)) %>%
  ggplot(aes(x = distance, y = avg_speed, color = Down)) +
  geom_point(aes(size = rushes), alpha = 0.4) +
  geom_smooth(aes(color = Down), se = FALSE, size = 2) +
  scale_color_viridis_d() +
  theme_reach() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  theme(legend.position = "bottom") +
  labs(x = "Yards to Sticks",
       y = "Average Speed",
       title = "How Down and Distance \n Impacts Running Back Speed",
       subtitle = "2020, speed is determined by handoff until first contact") +
  guides(size = FALSE)

```

Next we can take a look at yardline and box defenders

```{r plot 4, echo = FALSE, fig.align = "center", out.width= "100%"}
speed_yardline_box <- rushing_data_speed %>%
  filter(box_players %in% c(5, 6, 7, 8)) %>%
  mutate(yards_bin = round(yards_to_go / 5) * 5) %>%
  group_by(yards_bin, box_players) %>%
  summarize(rushes = n(),
            avg_speed = mean(avg_speed))

speed_yardline_box$box_players <- factor(speed_yardline_box$box_players, levels = c(5, 6, 7, 8))

speed_yardline_box %>%
  filter(avg_speed < 5.8) %>%
  mutate(`Box Defenders` = as.factor(box_players)) %>%
  ggplot(aes(x = yards_bin, y = avg_speed, color = `Box Defenders`)) +
  geom_point(aes(size = rushes), alpha = 0.4) +
  geom_smooth(aes(color = `Box Defenders`), se = FALSE, size = 2) +
  scale_color_viridis_d() +
  theme_reach() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  theme(legend.position = "bottom") +
  labs(x = "Yards From Endzone",
       y = "Average Speed",
       title = "How Yardline and Box Defenders Impacts \n Running Back Speed",
       subtitle = "2020, speed is determined by handoff until first contact") +
  guides(size = FALSE)
```

Most importantly, we can look at how different run concepts impact speed. 

```{r plot 5, echo = FALSE, fig.align = "center", out.width= "100%"}
rushing_data_speed %>%
  filter(concept_1 %in% c("Outside Zone", "Inside Zone", "Man", "Power", "Pull Lead", "Counter", "Draw", "Trap")) %>%
  group_by(concept_1) %>%
  mutate(speed = mean(avg_speed)) %>%
  ungroup() %>%
  mutate(concept_1 = fct_reorder(concept_1, speed)) %>%
  ggplot(aes(x = concept_1, y = avg_speed)) +
  geom_violin(aes(color = concept_1)) +
  geom_boxplot(aes(fill = concept_1)) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_reach() +
  labs(x = "Run Concept",
       y = "Speed Before Contact",
       title = "How Run Concept Affects a Running \n Back's Speed Through the Hole",
       subtitle = "2020, eight most common run types selected") +
  theme(axis.text.x = element_text(angle = -45, size = 12))
```

## Straight-Line Speed Over Expected (SSOE)

To adjust for all the factors listed earlier, we can use an xgboost model to created straight-line speed over expected, which is better at predicting yards before contact and more predictive than of next year's yards before contact (YBC).

```{r SSOE loading, echo = FALSE}
speed_projs <- read.csv("~/tracking-pff/speed_projs.csv")

round(summary(lm(ybc ~ speed_oe, data = speed_projs))$r.squared, 2)
round(summary(lm(ybc ~ avg_speed, data = speed_projs))$r.squared, 2)


speed_season_stats <- speed_projs %>%
  group_by(player, season, offense) %>%
  summarize(rushes = n(),
            avg_ybc = mean(ybc),
            exp_speed = mean(exp_speed),
            avg_speed = mean(avg_speed),
            avg_ssoe = mean(speed_oe)) %>%
  arrange(season) %>%
  group_by(player) %>%
  mutate(next_rushes = lead(rushes),
         next_ybc = lead(avg_ybc),
         next_exp_speed = lead(exp_speed),
         next_avg_speed = lead(avg_speed),
         next_ssoe = lead(avg_ssoe)) %>%
  filter(rushes >= 100) %>%
  filter(next_rushes >= 100 | is.na(next_rushes)) %>%
  left_join(teams_colors_logos, by = c("offense" = "team_abbr"))
```

```{r SSOE corr, echo = TRUE}
round(summary(lm(next_ybc ~ avg_ssoe, data = speed_season_stats))$r.squared, 2)
round(summary(lm(next_ybc ~ exp_speed, data = speed_season_stats))$r.squared, 2)
round(summary(lm(next_ybc ~ avg_speed, data = speed_season_stats))$r.squared, 2)
round(summary(lm(next_ybc ~ avg_ybc, data = speed_season_stats))$r.squared, 2)
```

Now that we've established what speed over expected can be used for, we can graph how it impacts yards before contact from 2018 to 2020. 

```{r plot 6, echo = FALSE, fig.align = "center", out.width= "100%"}
speed_oe_stats <- speed_projs %>%
  group_by(player) %>%
  summarize(rushes = n(), 
            avg_speed_oe = mean(speed_oe),
            avg_ybc = mean(ybc)) %>%
  filter(rushes >= 250) %>%
  arrange(-avg_speed_oe) %>%
  left_join(rb_colors, by = c("player"))

speed_oe_stats %>%
  ggplot(aes(x = avg_speed_oe, y = avg_ybc)) +
  geom_smooth(method = "lm", size = 1.5, color = "black", se = FALSE) +
  geom_hline(yintercept = mean(speed_season_stats$avg_ybc), linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = mean(speed_season_stats$avg_speed_oe), linetype = "dashed", alpha = 0.5) +
  geom_point(aes(fill = team_color, color = team_color2, size = rushes), shape = 21, alpha = 0.9) +
  ggrepel::geom_text_repel(aes(label = player), size = 3, box.padding = 0.3) +
  theme_reach() +
  scale_color_identity(aesthetics = c("fill", "color")) +
  labs(x = "Speed Over Expected",
       y = "Yards Before Contact",
       title = "How Speed Over Expected \n Impacts Yards Before Contact, 2018-2020",
       subtitle = "Minimum of 250 rushes") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6))
```

Looking at 2018 through 2020 is interesting, but focusing on just this past year is intriguing too

```{r plot 7, echo = FALSE, fig.align = "center", out.width= "100%"}

speed_season_stats %>%
  filter(season == 2020) %>%
  ggplot(aes(x = avg_ssoe, y = avg_ybc)) +
  geom_smooth(method = "lm", size = 1.5, color = "black", se = FALSE) +
  geom_hline(yintercept = mean(speed_season_stats$avg_ybc), linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = mean(speed_season_stats$avg_ssoe), linetype = "dashed", alpha = 0.5) +
  geom_point(aes(fill = team_color, color = team_color2, size = rushes), shape = 21, alpha = 0.9) +
  ggrepel::geom_text_repel(aes(label = player), size = 3, box.padding = 0.3) +
  theme_reach() +
  scale_color_identity(aesthetics = c("fill", "color")) +
  labs(x = "Speed Over Expected",
       y = "Yards Before Contact",
       title = "How Speed Over Expected \n Impacted Yards Before Contact in 2020",
       subtitle = "Minimum of 100 rushes") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6))
```

We can also look at it froma  team level

```{r plot 8, echo = FALSE, fig.align = "center", out.width= "100%"}

team_speed <- speed_projs %>%
  group_by(offense) %>%
  summarize(avg_exp_speed = mean(exp_speed),
            avg_speed = mean(avg_speed),
            avg_speed_oe = avg_speed - avg_exp_speed,
            avg_ybc = mean(ybc)) %>%
  left_join(teams_colors_logos, by = c("offense" = "team_abbr"))

team_speed %>%
  ggplot(aes(x = avg_exp_speed, y = avg_speed)) +
  geom_smooth(method = "lm", size = 1.5, color = "black", se = FALSE) +
  geom_hline(yintercept = mean(team_speed$avg_speed), alpha = 0.5, linetype = "dashed") +
  geom_vline(xintercept = mean(team_speed$avg_exp_speed), alpha = 0.5, linetype = "dashed") +
  geom_image(aes(image = team_logo_espn), asp = 16/9, size = 0.05) +
  theme_reach() +
  labs(x = "Average Expected Speed",
       y = "Average Actual Speed",
       title = "Actual and Expected Speed on Rushing Plays, 2020",
       subtitle = "Rushing speed determined by a \n running back's speed from handoff to first contact") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6))
```


